------------------------------------------------------------------------

------------------------------------------------------------------------

# Testing some stereotypes

```{r}
library(here)

snore <- read.table(
here("data", "snore.txt"),
header = TRUE
)
```

```{r}
head(snore)
```

## Exploring the snore data

```{r}
str(snore)
summary(snore)
names(snore)
```

The dataset contains anthropometric measurements (age, weight, height) and categorical variables describing sex, smoking status, alcohol consumption and snoring status.

Statistical comparisons were performed to evaluate whether common stereotypes are supported by the data.

## Are snorers fatter?

```{r s23-boxplot-snore}
boxplot( 
  weight ~ snore,
  data = snore,
  col = c("lightblue","salmon"),
  xlab = "Snoring",
  ylab = "Weight (kg)",
  main = "Weight according to snoring status" 
  )
```

```{r}
wilcox.test(weight ~ snore, data = snore)
```

### conclusion

The data do not support the stereotype that snorers are heavier than non-snorers. A Wilcoxon rank-sum test showed no significant difference in weight between the two groups (W = 1154, p = 0.908).


## Do snorers drink more alcohol?
Let's compare alcohol consumption between snorers and non-snorers.

```{r}
boxplot(
alcool ~ snore,
data = snore,
col = c("lightblue", "salmon"),
xlab = "Snoring",
ylab = "Alcohol consumption",
main = "Alcohol consumption by snoring status"
)
```
```{r}
wilcox.test(alcool ~ snore, data = snore)
```

### conclusion

Alcohol consumption was compared between snorers and non-snorers using a Wilcoxon rank-sum test. The test showed a statistically significant difference between the two groups (W = 788, p = 0.009). Therefore, the data support the stereotype that snorers consume more alcohol than non-snorers.

## Do snorers smoke more?
```{r}
table(snore$snore, snore$tabac)
```
```{r}
chisq.test(table(snore$snore, snore$tabac))
```
### conclusion
The test showed no significant association between smoking and snoring status (χ² = 0.689, df = 1, p = 0.407). Therefore, the data do not support the stereotype that snorers smoke more than non-snorers.


## Are men fatter?
```{r}
boxplot(
weight ~ sex,
data = snore,
col = c("pink", "lightblue"),
xlab = "Sex",
ylab = "Weight (kg)",
main = "Weight according to sex"
)
```

```{r}
wilcox.test(weight ~ sex, data = snore)
```

### conclusion
The test showed no significant difference in weight between the two groups (W = 893.5, p = 0.729). Therefore, the data do not support the stereotype that men are fatter than women.

