------------------------------------------------------------------------

------------------------------------------------------------------------

# Testing some stereotypes

```{r}
library(here)

snore <- read.table(
here("data", "snore.txt"),
header = TRUE
)
```

```{r}
head(snore)
```

## Exploring the snore data

```{r}
str(snore)
summary(snore)
names(snore)
```

The dataset contains anthropometric measurements (age, weight, height) and categorical variables describing sex, smoking status, alcohol consumption and snoring status.

Statistical comparisons were performed to evaluate whether common stereotypes are supported by the data.

## Are snorers fatter?

```{r s23-boxplot-snore}
boxplot( 
  weight ~ snore,
  data = snore,
  col = c("lightblue","salmon"),
  xlab = "Snoring",
  ylab = "Weight (kg)",
  main = "Weight according to snoring status" 
  )
```

```{r}
wilcox.test(weight ~ snore, data = snore)
```

### conclusion

The data do not support the stereotype that snorers are heavier than non-snorers. A Wilcoxon rank-sum test showed no significant difference in weight between the two groups (W = 1154, p = 0.908).


