------------------------------------------------------------------------

------------------------------------------------------------------------

# Effect of treatment over time

## The data

```{r s22-load-data}
library(tidyverse)
library(here)

prepost <- read.csv(here("data", "PrePost.csv"))
str(prepost)
summary(prepost)
table(prepost$time)
```

The file `PrePost.csv` contains 16 observations and 2 variables: a performance measure (`perf`) and the time of measurement (`time`), with 8 measurements before and 8 measurements after the rehabilitation program.

No variable identifies the participants, so the file itself does not tell us whether the same individuals were measured before and after.

| Variable | Description | Unit | Possible values | Expected change |
|---------------|---------------|---------------|---------------|---------------|
| `perf` | Physical performance measured before and after the rehabilitation program | Not specified | Positive numerical values; observed range: 125–148 | Expected to increase after rehabilitation |
| `time` | Time of measurement relative to the rehabilitation program | Not applicable | `"Before"`, `"After"` | / |

## The analysis

### 1. Does the treatment have an effect?

The objective is to determine whether the rehabilitation program has an effect on physical performance. Because rehabilitation is expected to improve the patients' performance, we expect the performance values to be higher **after** the rehabilitation program than **before**. The comparison of interest is therefore: **Before vs After rehabilitation**.

```{r s22-desc}
# descriptive statistics for each time point
prepost %>%
  mutate(time = factor(time, levels = c("Before", "After"))) %>%
  group_by(time) %>%
  summarise(
    n      = n(),
    mean   = mean(perf),
    sd     = sd(perf),
    median = median(perf),
    IQR    = IQR(perf),
    min    = min(perf),
    max    = max(perf)
  )
  
```

```{r s22-diff}
# difference between after and before
mean(prepost$perf[prepost$time == "After"]) -
  mean(prepost$perf[prepost$time == "Before"])

median(prepost$perf[prepost$time == "After"]) -
  median(prepost$perf[prepost$time == "Before"])
```

Descriptively, performance was slightly higher after rehabilitation than before, both in mean (137.6 ± 7.0 vs 135.1 ± 6.2, i.e. +2.5 units) and in median (137.5 \[IQR = 9.25\] vs 135.5 \[IQR = 7.00\], i.e. +2 units). Mean ± SD is appropriate only if the data are normally distributed; otherwise, the median and IQR should be preferred. Normality is checked in question 3. In both cases, the difference is small compared with the variability between individuals, and the two distributions largely overlap (range 125–145 before vs 127–148 after). Whether this difference is statistically significant is tested in question 5.

### 2. What comparison are you going to make?

We compare physical performance **before** vs **after** the rehabilitation program (8 observations in each group).

Ideally, a before–after study uses a **paired comparison**, because the same patients are measured twice. The analysis is then based on the difference computed for each participant:

$$\text{Difference} = \text{After} - \text{Before}$$

However, the dataset does not contain any participant identifier, so it is impossible to know which "Before" value corresponds to which "After" value.

```{r s22-sorted}
before <- prepost$perf[prepost$time == "Before"]
after  <- prepost$perf[prepost$time == "After"]

before
after

# are the values sorted in increasing order within each group?
!is.unsorted(before)
!is.unsorted(after)
```

Moreover, the values are sorted in increasing order within each group. Pairing them by row order would match the lowest "Before" value with the lowest "After" value, and so on, which would artificially create positive differences and bias a paired test.

For these reasons, the two sets of observations are treated as **independent groups**. This is a conservative choice: if the data were truly paired, a paired test would be more powerful, so we may miss a real effect.

### 3. Which test will you use? Parametric or non-parametric?

**Hypotheses** (two-sided test):

- H0: performance does not differ between before and after rehabilitation;
- H1: performance differs between before and after rehabilitation.

Although an increase is expected, a two-sided test is used, as it is more conservative and is the standard choice.

The choice of test depends on whether the observations are paired and on the distribution of the data:

| Comparison       | Parametric test | Non-parametric test                   |
|------------------|-----------------|---------------------------------------|
| Paired data      | Paired t-test   | Wilcoxon signed-rank test             |
| Independent data | Welch t-test    | Mann–Whitney / Wilcoxon rank-sum test |

Since the groups are treated as independent (question 2), the choice is between the Welch t-test and the Wilcoxon rank-sum test. Normality is checked in each group with the Shapiro–Wilk test:

```{r s22-normality}
shapiro.test(before)
shapiro.test(after)
```

The Shapiro–Wilk test did not detect any deviation from normality, neither before (W = 0.997, p = 0.9999) nor after rehabilitation (W = 0.983, p = 0.977). However, with only 8 observations per group, this test has limited power. Since no deviation from normality was detected, the data are summarised with mean ± SD and a **Welch t-test** is used. The result is confirmed with its non-parametric equivalent, the **Wilcoxon rank-sum test**, as a robustness check.
