------------------------------------------------------------------------

------------------------------------------------------------------------

# Effect of treatment over time

## The data

```{r s22-load-data}
library(tidyverse)
library(here)

prepost <- read.csv(here("data", "PrePost.csv"))
str(prepost)
summary(prepost)
table(prepost$time)
```

The file `PrePost.csv` contains 16 observations and 2 variables: a performance measure (`perf`) and the time of measurement (`time`), with 8 measurements before and 8 measurements after the rehabilitation program.

No variable identifies the participants, so the file itself does not tell us whether the same individuals were measured before and after.

| Variable | Description | Unit | Possible values | Expected change |
|---------------|---------------|---------------|---------------|---------------|
| `perf` | Physical performance measured before and after the rehabilitation program | Not specified | Positive numerical values; observed range: 125–148 | Expected to increase after rehabilitation |
| `time` | Time of measurement relative to the rehabilitation program | Not applicable | `"Before"`, `"After"` | / |

## The analysis

### 1. Does the treatment have an effect?

The objective is to determine whether the rehabilitation program has an effect on physical performance.

Because rehabilitation is expected to improve the patients' performance, we expect the performance values to be higher **after** the rehabilitation program than **before**.

Therefore, the comparison of interest is: **Before vs After rehabilitation.**

### 2. Which type of comparison should be used?

Ideally, a before-after study would use a **paired comparison**, because the same patient would be measured before and after rehabilitation.

In a paired analysis, the difference would be calculated for each participant:

$$\text{Difference} = \text{After} - \text{Before}$$

However, the dataset does not contain any participant identifier. Therefore, it is impossible to determine from the file whether each `"Before"` observation corresponds to a specific `"After"` observation.

For this reason, the two sets of observations will be treated as **independent groups** for the statistical analysis.

### 3. Parametric or non-parametric test?

The choice of statistical test depends on the distribution of the data and on whether the observations are paired or independent.

| Comparison       | Parametric test | Non-parametric test                   |
|------------------|-----------------|---------------------------------------|
| Paired data      | Paired t-test   | Wilcoxon signed-rank test             |
| Independent data | Welch t-test    | Mann–Whitney / Wilcoxon rank-sum test |

There are only **8 observations per group**, so the sample size is very small.

A normality test can be performed, but with such a small sample it has limited power to detect deviations from normality.

Because the pairing between observations cannot be established from the dataset, we will consider an **independent comparison**.

Before choosing between the Welch t-test and the Wilcoxon rank-sum test, the distribution of the data will be explored graphically and with descriptive statistics.

### 4. Which graph should be used?

A boxplot can be used to compare performance before and after rehabilitation.

Because there are only 8 observations in each group, the individual observations should also be displayed on the graph.

This makes it possible to see both the overall distribution and each individual value.

### 5. How should the result be reported?

The final conclusion will depend on the p-value obtained from the statistical test.

If:

$p < 0.05$

we will reject the null hypothesis and conclude that there is evidence of a difference in performance before and after rehabilitation.

If:

$p \geq 0.05$

we will not reject the null hypothesis, meaning that the data do not provide sufficient evidence of a significant difference.

The final result can be reported in the following form:

*"Performance was \[significantly / not significantly\] different after rehabilitation compared with before rehabilitation (median: X before vs Y after; \[name of test\], p = X)."*

## The analysis

### Descriptive statistics

```{r s22-desc}
prepost %>%
  group_by(time) %>%
  summarise(n = n(), mean = mean(perf), sd = sd(perf),
            median = median(perf), IQR = IQR(perf))
```

### Normality

```{r s22-normality}
shapiro.test(prepost$perf[prepost$time == "Before"])
shapiro.test(prepost$perf[prepost$time == "After"])
```

### Graph

```{r s22-boxplot}
prepost %>%
  mutate(time = factor(time, levels = c("Before", "After"))) %>%
  ggplot(aes(x = time, y = perf, fill = time)) +
    geom_boxplot(alpha = 0.5) +
    geom_jitter(width = 0.1) +
    labs(x = "", y = "Performance") +
    theme_minimal()
```

### Statistical tests

```{r s22-tests}
before <- prepost$perf[prepost$time == "Before"]
after  <- prepost$perf[prepost$time == "After"]

# unpaired (independent groups)
wilcox.test(after, before, paired = FALSE)

# paired (same patients, in the same order)
wilcox.test(after, before, paired = TRUE)
```
